\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-08 analysis/paper/paper.tex

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Time matching/risk set sampling (including Cox partial likelihood)
  eliminates the baseline hazard from the likelihood expression for the
  hazard ratios
\item
  If, however, the absolute risks are of interest, they have to be
  recovered using the semi-parametric Breslow estimator
\item
  Alternative approaches for fitting flexible hazard models for
  estimating absolute risks, not requiring this two-step approach? Yes!
  \citep{hanley2009fitting}
\end{enumerate}

@@

@@

\hypertarget{case-study-3the-cancer-genome-atlas}{%
\section{Case study 3--The Cancer Genome
Atlas}\label{case-study-3the-cancer-genome-atlas}}

\begin{itemize}
\tightlist
\item
  Glmnet and TCGA data
\end{itemize}

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    796    760    730    728
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>         
#> 365 0.27
\end{CodeOutput}

@@

\begin{CodeOutput}
#> [1] 0.18
\end{CodeOutput}
\end{CodeChunk}

@@

This report was generated on 2019-12-08 11:49:58 using the following
computational environment and dependencies:

@@

@@

@@

@@

@@

#>  Matrix        1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  

@@

\begin{CodeOutput}
#> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [f601653] 2019-12-08: Merge branch 'review-max' of https://github.com/sahirbhatnagar/cbpaper into review-max
\end{CodeOutput}
\end{CodeChunk}

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  hazard ratios
\item
  If, however, the absolute risks are of interest, they have to be
  recovered using the semi-parametric Breslow estimator
\item
  Alternative approaches for fitting flexible hazard models for
  estimating absolute risks, not requiring this two-step approach? Yes!
  \citep{hanley2009fitting}
\end{enumerate}

@@

\hypertarget{literature-review}{%
\section{Literature review}\label{literature-review}}

To conduct our review on current methods available in survival analysis,
a scraper was coded to examine all packages in the survival task view on
CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
of which are not relevant to the problems casebase addresses. To filter
these packages, a scraper was used to search for competing risks,
cumulative incidence functions, non- proportional hazards and
penalization. This reduced our set of 259 packages down to 60. From
here, we manually examined each package to determine which ones are
comparable to casebase, with either a different approach to the same
problems or future features we would like to add to casebase.

These packages can belong to three main categories: non-proportional
hazards, regularization and competing risks. Besides these three main
categories, there are extra features of interest as well. Table
\ref{tab:surv-pkgs} summarizes the findings.

Many packages exist that implement semi-parametric Cox models. Of those,
\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package} include penalization, but they implement
the extended cox model permitting intuitive non proportional hazard
modeling while the previous four do not. In terms of penalization, the
benefits of \pkg{casebase} comes from its flexibility. Linear models
have a wide tool set, considerably larger than those made for Cox
regression. \pkg{casebase} permits many regularization tools developed
for linear regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in
a handful of packages. Of interest are \pkg{CFC}
\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
\pkg{survival} contain implementations for users to supply any
distribution they would like \citeyearpar{survival-package},
\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
and \pkg{rstpm2} can use splines, permitting a flexible fit over time
\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
\pkg{casebase} is limited to three distributions at the moment, it also
uses splines allowing for distributions that may not be consistent over
survival time.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
differentiating factor between these packages and \pkg{casebase} is that
\pkg{casebase} handles competing risks through multiple logistic
regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
handle interval censoring, whereas \pkg{casebase} currently cannot.
Interval censoring is a simple feature to implement and is a potential
improvement to add to \pkg{casebase}.

Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
\pkg{survival} have implementations for cumulative incidence
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
\pkg{casebase} is parametric, the baseline hazard is known and requires
integration to retrieve the cumulative incidence. This is similar to the
approaches of \pkg{CFC} and \pkg{flexsurv}
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
other packages use semi-parametric approaches to retrieve the baseline
after fitting \citeyearpar{scheike2014estimating},
\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
for their respective parametric and semi-parametric model
implementations \citeyearpar{clements_liu},
\citeyearpar{survival-package}.

\begin{table}[ht]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{ccccccccc}
\hline
\textbf{Package}        & \textbf{Competing Risks} & \textbf{Non-proportional} & \textbf{Penalization} & \textbf{Splines} & \textbf{Parametric} & \textbf{Semi-parametric} & \textbf{Interval/left Censoring} & \textbf{Absolute Risk}           \\
\hline
\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\
\textbf{CFC} \cite{sharabiani_mahani_2019}            & x                        & x                         &                       &                  & x                   &                          &                                  & x                                \\
\textbf{coxRidge} \cite{perperoglou}       &                          & x                         & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{crrp} \cite{fu}           & x                        &                           & x                     &                  &                     &                          &                                  &                                  \\
\textbf{fastcox} \cite{yi_zou}        &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{Flexrsurv} \cite{flexrsurv}      &                          & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?               \\
\textbf{Flexsurv} \cite{flexsurv}       & x                        & x                         &                       & x                & x                   &                          &                                  & Cumulative hazard?               \\
\textbf{glmnet} \cite{regpathcox}         &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{glmpath} \cite{park_hastie}       &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{mets} \cite{scheike2014estimating}           & x                        &                           &                       & x                &                     & x                        &                                  & x                                \\
\textbf{penalized} \cite{goeman_meijer2019}      &                          &                           & x                     &                  &                     & x                        &                                  &                                  \\
\textbf{RiskRegression} \cite{gerds_blanche} &                          &                           & x                     &                  &                     & x                        &                                  & x                                \\
\textbf{Rstpm2} \cite{clements_liu}         &                          & x                         & x                     & x                & x                   & x                        & x                                & Cumm haz                         \\
\textbf{SmoothHazard} \cite{smoothHazard}   &                          & x                         &                       & x                & x                   &                          & x                            &                                      \\
\textbf{Survival} \cite{survival-package}       & x                        & x                         &                       &                  & x                   & x                        & x                                & x                               
\end{tabular}%
}
\caption{Different features of interest in various survival packages.}
\label{tab:surv-pkgs}
\end{table}

@@

\hypertarget{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}{%
\section{Case study 3--Study to Understand Prognoses Preferences
Outcomes and Risks of
Treatment}\label{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}}

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT
dataset tracks death for individuals who are considered seriously ill
within a hospital. Imputation was conducted using the default methods in
the mice package in R, available on CRAN. Before imputation, there was a
total of 1000 individuals and 35 variables. After imputation, we have
474 observations where 67.5\% of individuals in the study experience the
event of interest. The SUPPORT dataset will be used to demonstrate
regularization within casebase (REF), while comparing their absolute
risk preditions for a new individual. A description of each variable can
be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown of
each categorical variable in Table\textasciitilde{}\ref{tab:support2}.
The data was downloaded from the main author's website, and it is also
available as part of the casebase package.

\begin{table}[ht]
\centering
\begin{tabular}{ccccc}
\hline
\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
\hline
age           & Age                                      &                 & double           & 0            \\
death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
sex           &                                          & 2               & integer          & 0            \\
hospdead      & Death in Hospital                        &                 & double           & 0            \\
slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
d.time        & Days of Follow-Up                        &                 & double           & 0            \\
dzgroup       &                                          & 8               & integer          & 0            \\
dzclass       &                                          & 4               & integer          & 0            \\
num.co        & number of comorbidities                  &                 & double           & 0            \\
edu           & Years of Education                       &                 & double           & 202          \\
income        &                                          & 4               & integer          & 349          \\
scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
charges       & Hospital Charges                         &                 & double           & 25           \\
totcst        & Total RCC cost                           &                 & double           & 105          \\
totmcst       & Total micro-cost                         &                 & double           & 372          \\
avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
race          &                                          & 5               & integer          & 5            \\
meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
temp          & Temperature (celcius) Day 3              &                 & double           & 0            \\
pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
bili          & Bilirubin Day 3                          &                 & double           & 297          \\
crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
sod           & Serum sodium Day 3                       &                 & double           & 0            \\
ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
glucose       & Glucose Day 3                            &                 & double           & 470          \\
bun           & BUN Day 3                                &                 & double           & 455          \\
urine         & Urine Output Day 3                       &                 & double           & 517          \\
adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
sfdm2         &                                          & 5               & integer          & 159          \\
adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
\end{tabular}
\caption{A description of each variable in the SUPPORT dataset, taken from (REF).}
\label{tab:support1}
\end{table}

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
\label{tab:support2}
\end{table}

Hospital death as a covariate was removed, as this is directly
informative of death. The glmnet package(REF) on CRAN was used to
introduce elastic net, lasso and ridge regressions to casebase. To
visualize the effect of imputation on incidence density, we can refer to
the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and
post-imputation population time plot in Figure \ref{fig:ptPostI}. The
incidence density is consistent between imputed and complete datasets.

\begin{CodeInput}
R> #demonstrating incidence density before imputation
R> plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

\begin{CodeInput}
R> #demonstrating incidence density after imputation
R> plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

Our main objective is to compute the absolute risk of death for a given
set of covariates, using regularization when fitting our model. First,
we fit a smooth hazard to the data; for the sake of this example, we
opted for a linear term for time. As casebase makes use of the glmnet
package, we interact with the fitsmoothhazard.fit function using a
matrix interface, where y contains the time and event variables and x
contains all other variables we would like to include in the model. For
the SUPPORT dataset, all factor variables were converted into dummy
variables. Here, we used lasso penalization by setting alpha to 1.

\begin{CodeInput}
R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
R> #to determine the hazard function
R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
R>                              formula_time = ~d.time,alpha=1,ratio=100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model and use the absoluteRisk function to integrate
and retrieve our absolute risk function.

\begin{CodeInput}
R> #Estimating the absolute risk curve using the newData parameter.
R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
R>                               s="lambda.1se",method=c("numerical"))
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of cox regression,
and a kaplan-meier survival curve. The regularized cox regression
absolute risk curve required two extra steps to retrive. Coxnet from the
glmnet package has tools to fit a regularized cox model, but requires
that a survival object skeleton of the selected variables, so that the
survival package tools can be used to retrive the absolute risk. We
handle this by fitting a regularized cox model with the glmnet package,
and a cox model from the survival package with the selected variables
from the regularized fit. This second model serves as a skeleton that
permits the use of survival package functions. The coefficients from the
regularized model will replace the coeficients in our skeleton. the
resulting model will have the correct coefficients when integrating, but
will have an incorrect standard error. For the purposes of this case
study, we are only interested in the absolute risk curve itself and not
the standard error.

\begin{CodeInput}
R> #Create u and xCox, which will be used when fitting Cox with glmnet.
R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
R> #hazard for cox using glmnet
R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
R> #convergence demonstrated in plot
R> plot(coxglmFit)
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
R> #creating a new dataset that only contains the covariates chosen through glmnet.
R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
R> #fitting a cox model using regular estimation, however we will not keep it.
R> #this is used more as an object place holder.
R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
R> #Doing so makes it so that everything is invalid aside from the coefficients.
R> #In this case, all we need to estimate the absolute risk is the coefficients.
R> #Std. error would be incorrect here, if we were to draw error bars.
R> coxFit$coefficients<-nonzero_coef_cox@x
R> #Fitting absolute risk curve for cox+glmnet
R> newDataCox=newData[nonzero_covariate_cox$X1-1]
R> 
R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
\end{CodeInput}
\end{CodeChunk}

We used the survival package to calculate the kaplan-meier absolute risk
function.

\begin{CodeInput}
R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
R> # (Kaplan-Meier risk curve)
R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
\end{CodeInput}
\end{CodeChunk}

Now that our three absolute risk functions have been calculated, we
compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
decrease the absolute risk by the end of the study, in comparison to
kaplan-meier.

\begin{CodeInput}
R> # A plot to compare all three Absolute risk (Commulative Incidence)
R> plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
R> lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
R> legend("bottom", 
R>        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
R>        col = c("black","red","Blue"),
R>        lty = c(1, 1, 1), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    798    759    737    708
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeOutput}
#>         
#> 365 0.32
\end{CodeOutput}

@@

\begin{CodeOutput}
#> [1] 0.24
\end{CodeOutput}
\end{CodeChunk}

@@

This report was generated on 2019-12-08 12:31:41 using the following
computational environment and dependencies:

@@

#>  codetools     0.2-16     2018-12-24 [4] CRAN (R 3.5.2)                  

@@

#>  foreach     * 1.4.4      2017-12-12 [1] CRAN (R 3.5.0)                  

@@

#>  glmnet      * 2.0-18     2019-05-20 [1] CRAN (R 3.6.1)                  

@@

#>  iterators     1.0.9      2017-12-12 [1] CRAN (R 3.5.0)                  

@@

#>  Matrix      * 1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  

@@

\begin{CodeOutput}
#> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [69663bb] 2019-12-08: Merge branch 'review-max' of https://github.com/sahirbhatnagar/cbpaper into review-max
\end{CodeOutput}
\end{CodeChunk}
